---
title: "Practical Machine Learning course project"
author: "by Jos&eacute; Alberto Hern&aacute;ndez"
date: "May 22nd, 2016"
output:
  html_document:
    css: css/style.css
    keep_md: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
geometry: margin=0.8in
---

```{r opts, echo=FALSE}
knitr::opts_chunk$set(
  fig.path = "figure/",
  cache = TRUE
)
```

# Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.  

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise.  

```{r row-data, echo=FALSE}
# Training data
# https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
# Test data
# https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

## Data

The training data for this project are available [here](`r trainUrl`)

The test data are available [here](`r testUrl`)

# R Packages
We validate the R packages need if they are installed, If not we install them
```{r validate-packages, results='hide'}
packages <- c("caret","rpart","rpart.plot","RColorBrewer","rattle","randomForest","knitr","RCurl")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))  
}
```

Load the R packages
```{r load-packages, warning=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
library(RCurl)
```

# Getting the row data
```{r get-data}
# Training data file
trainFile <- "./data/pml-training.csv"

# Test data file
testFile <- "./data/pml-testing.csv"

if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}
```  

### Read the Data
After downloading the data from the data source, we can read the two csv files into two data frames.  
```{r read-data}
# Reading the csv files to dataframes
training <- read.csv(trainFile, na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(testFile, na.strings=c("NA","#DIV/0!",""))
```

The training data set contains `r nrow(training)` observations and `r ncol(training)` variables, while the testing data set contains `r nrow(testing)` observations and the same `r ncol(testing)` variables. 

# Create data partitions
```{r data-partition}
set.seed(12345)

# Partition the training data for Cross Validation
inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
```

The training partition contains `r nrow(myTraining)` observations, while the testing partion contains `r nrow(myTesting)` observations, with a total `r nrow(myTraining) + nrow(myTesting)` as the training data set. 

# Cleaning the data
```{r clean-data}
## Remove NearZeroVariance variables
nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,nzv$nzv==FALSE]

## Remove the first column of the myTraining data set
myTraining <- myTraining[c(-1)]

## Clean variables with more than 60% NA
trainingV3 <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) == 1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}

# Set back to the original variable name
myTraining <- trainingV3
rm(trainingV3)
```

## Transform the myTesting and testing data sets
``` {r tranform-data}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])  # remove the classe column
myTesting <- myTesting[clean1]         # allow only variables in myTesting that are also in myTraining
testing <- testing[clean2]             # allow only variables in testing that are also in myTraining
```

The myTesting data set has `r nrow(myTesting)` rows with `r ncol(myTesting)` cols
The testing data set has `r nrow(testing)` rows with `r ncol(testing)` cols

## Coerce the data into the same type
```{r coerce-data}
for (i in 1:length(testing) ) {
    for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}

## To get the same class between testing and myTraining
testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```

# Data modeling and Predictions
## Prediction with Decision Trees
``` {r prediction-trees}
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)

predictionsA1 <- predict(modFitA1, myTesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, myTesting$classe)
cmtree

plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

## Prediction with Random Forests
``` {r prediction-forests}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data=myTraining)
predictionB1 <- predict(modFitB1, myTesting, type = "class")
cmrf <- confusionMatrix(predictionB1, myTesting$classe)
cmrf

plot(modFitB1)

plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```

## Prediction with Generalized Boosted Regression
``` {r prediction-gbr}
set.seed(12345)
fitControl <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 1)

gbmFit1 <- train(classe ~ ., data=myTraining, method = "gbm",
                 trControl = fitControl,
                 verbose = FALSE)

gbmFinMod1 <- gbmFit1$finalModel

gbmPredTest <- predict(gbmFit1, newdata=myTesting)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, myTesting$classe)
gbmAccuracyTest

plot(gbmFit1, ylim=c(0.9, 1))
```

## Predicting Results on the Test Data
Now, we apply the model to the original testing data set downloaded from the data source.
``` {r prediction-test-data}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
```
``` {r write-results, echo=FALSE}
# Write the results to a text file for submission
pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}
```